---
title: "Data portals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_portals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(httptest2)
start_vignette('data_portals')
```

```{r setup}
library(hydrogauge)
library(ggplot2)
```

Most examples use Victorian water data for hydstra or BOM for KiWIS, but this should work wherever we have a link to a hydstra or Kiwis API. Many sites are listed by [Kisters](https://resources.kisters.com.au/public/kisters-web-publishing/), though not all make it easy to find their path. 

At the most basic, we can use a raw https path, demonstrated with Victoria,


```{r}
ds_http <- get_datasources_by_site(portal = 'https://data.water.vic.gov.au/WMIS/cgi/webservice.exe?', 
                              site_list = c('233217', '405328', '405331', '405837'))
```

```{r rows.print = 15}
ds_http
```

## Hydstra

In theory, we should be able to call the hydstra functions at any Hydstra source. The catch is, finding the API path is not always straightforward. For example, this *should* work for [Washington Department of Ecology](https://apps.ecology.wa.gov/continuousflowandwq/), [California department of water resources](https://wdl.water.ca.gov/waterdatalibrary/), and [Western Australia](https://kumina.water.wa.gov.au/waterinformation/telem/stage.cfm), but none of those addresses work. Either those orgs don't publish their APIs, or there are undiscovered '.../webservice.exe?' paths there waiting to be discovered and used.  

### Australian states

In addition to Victoria, New South Wales and Queensland have public portals and are available by name as `portal` arguments. To demonstrate, we can use `get_datasources_by_site()` to see what datasources are available (note the major differences between the states):

```{r}
nsw_ds <- get_datasources_by_site(portal = 'NSW', 
                                  site_list = c("422028", "410007"))
```

```{r}
nsw_ds
```

```{r}
qld_ds <- get_datasources_by_site(portal = 'QLD', 
                                  site_list = c("423203A", "424201A"))
```

```{r}
qld_ds
```

#### Traces and plots

Now we can `get_ts_traces()` for `A`, just to keep things consistent for NSW:

```{r}
nsw_ts_days <- get_ts_traces(portal = 'NSW', 
                         site_list = c("422028", "410007"),
                         datasource = "A", 
                         var_list = c("100", "141", "450"),
                         start_time = 20200101,
                         end_time = 20201231,
                         interval = "day",
                         data_type = "mean",
                         multiplier = 1,
                         returnformat = 'df')

```

```{r}
ggplot(nsw_ts_days, aes(x = time, y = value, color = site_short_name)) + 
  facet_wrap(~variable_short_name, scales = 'free') + 
  geom_line()
```

and QLD:

```{r}
qld_ts_days <- get_ts_traces(portal = 'QLD', 
                         site_list = c("423203A", "424201A"),
                         datasource = "A", 
                         var_list = c("100", "141", "450"),
                         start_time = 20200101,
                         end_time = 20201231,
                         interval = "day",
                         data_type = "mean",
                         multiplier = 1,
                         returnformat = 'df')

```

```{r}
ggplot(qld_ts_days, aes(x = time, y = value, color = site_short_name)) + 
  facet_wrap(~variable_short_name, scales = 'free') + 
  geom_line()
```

```{r, include=FALSE}
# For httptest2 mocking
end_vignette()
```
