---
title: "hydstra wrapper"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(httptest2)
start_vignette('hydstra_wrapper')
```

```{r}
library(hydrogauge)
library(ggplot2)
```

The `fetch_hydstra_timeseries()` function wraps `get_variable_list()` and `get_ts_traces()`, which allows some extra functionality and smoother workflows but also introduces some danger and sometimes inefficiency. Some argument names have been changed compared to `get_ts_traces()` (which uses the Kisters names nearly verbatim) for more clarity and to move towards a unified interface across the KiWIS and hydstra API styles.

This function allows requests for the full period of record by passing `'all'` for the `start_time` and `end_time` (or the earliest and latest possible dates, respectively). The downside of this approach is that the calls to the API are inefficient (each row is called separately), though this is also necessarily the case if asking for the period of record manually for gauges with different periods. It will be possible to do some inferences and combinations here, but it has not been high priority. 

This function is most useful when we want to pull the period of record of the same variable for a set of gauges. For example, we might want to pull discharge for the period of record, which we can do by passing `'all'` to `start_time` and `end_time`, yielding \@ref(fig:discharge-period).

## Period of record

I'll use the same set of sites as in the [core hydstra demonstration](hydstra_traces_workflow.Rmd), chosen to capture a range of periods of records and available variables.

-   The Upper Steavenson (405328) only has flow

-   Barwon (233217) has many variables, but their start dates differ

-   Taggerty (405331) is no longer in operation- ran 2010-2013

-   Marysville golf course (405837) is only rainfall

```{r}
barwon <- '233217'
steavenson <- '405328'
taggerty <- '405331'
golf <- '405837'
```

```{r}
discharge_record <- fetch_hydstra_timeseries(portal = 'vic', 
                                       gauge = c(barwon, steavenson, taggerty),
                                       var_list = '141',
                                       start_time = 'all',
                                       end_time = 'all',
                                       timeunit = 'day',
                                       statistic = 'mean')
```

```{r discharge-period, fig.cap = "Discharge for the period of record for three gauges."}
#| label: fig-discharge-period
#| fig-cap: Discharge for the period of record for three gauges.
discharge_record |> 
  ggplot(aes(x = time, y = value, color = site_short_name)) +
  geom_line() +
  facet_grid(site_short_name~., scales = 'free_y') +
  labs(y = unique(discharge_record$variable_short_name))
```

## Multiple data types

We can also pull data for all available variables by passing `'all'` to `var_list`. *DANGER: if `var_type = 'all'`, the same statistic will be applied to all variables.* Calling `'all'` for the `start_time` and `'end_time'` *will* give each variable a different period of record if they differ; the times are found from each row returned by get_variable_list(). 

For the sake of demonstration, we make the bad choice here of getting all the data, summarised in @tbl-all. This throws a warning because it's a bad idea in general.

```{r}
all_vars_fullperiod <- fetch_hydstra_timeseries(portal = 'vic', 
                                       gauge = c(barwon, golf),
                                       var_list = 'all',
                                       start_time = 'all',
                                       end_time = 'all',
                                       timeunit = 'day',
                                       statistic = 'mean')
```
```{r}
#| label: tbl-all
#| tbl-cap: Number of records for the period of record pulled by code above. Note that the Rainfall measurements will be daily means, which is not appropriate.
#| 
all_vars_fullperiod |> 
  dplyr::summarise(n_records = dplyr::n(), 
                   .by = c(site_short_name, variable_short_name, statistic)) |> 
  knitr::kable()
```

We *can* request different statistics for different variables if the variables are passed in as a vector of arguments to `var_list`, though then you have to know what they are. In that case, the `statistic` argument should be a vector of matched length to `var_list`.

Let's ask for the period of record for daily mean discharge, total daily rainfall, and maximum daily temperature in the Barwon (233217), which returns different statistics for each variable (@tbl-diffstats).

```{r}
different_statistics <- fetch_hydstra_timeseries(portal = 'vic', 
                                       gauge = c(barwon, golf),
                                       var_list = c('141', '10', '450'),
                                       start_time = 'all',
                                       end_time = 'all',
                                       timeunit = 'day',
                                       statistic = c('mean', 'tot', 'max'))
```

```{r}
#| label: tbl-diffstats
#| tbl-cap: Number of records for the period of record pulled by code above. By using a vector of statistics matching the vector of var_list, we can use different statistics for each variable.
#| 
different_statistics |> 
  dplyr::summarise(n_records = dplyr::n(), 
                   .by = c(site_short_name, variable_short_name, statistic)) |> 
  knitr::kable()
```

## Regex selection

We can also use the `variable` and `unit` arguments instead of `var_list` to search for variables by name, as in fetch_kiwis_timeseries(). This is very experimental, moving towards a unified wrapper. We can use this to recapitulate the pull of discharge for the gauges (\@ref(fig:discharge-byname)).

```{r}
by_name <- fetch_hydstra_timeseries(portal = 'vic', 
                                       gauge = c(barwon, steavenson, taggerty),
                                       variable = 'discharge',
                                       unit = 'ML/d',
                                       start_time = 'all',
                                       end_time = 'all',
                                       timeunit = 'day',
                                       statistic = 'mean')
```

```{r discharge-byname, fig.cap = "Discharge for the period of record for three gauges, obtained by name."}
#| label: fig-discharge-byname
#| fig-cap: Discharge for the period of record for three gauges, obtained by name.
by_name |> 
  ggplot(aes(x = time, y = value, color = site_short_name)) +
  geom_line() +
  facet_grid(site_short_name~., scales = 'free_y') +
  labs(y = unique(by_name$variable_short_name))
```

## Large requests

*Note: with big pulls, it can be useful to use the bare `get_variable_list()` and `get_ts_traces()` approach, or at least a manual check of `get_variable_list()`.* In my experience, there are often errors with some gauges or other issues that mean clean pulls need some troubleshooting of the variable availability etc. It is often easiest to find and solve problems at the low-level API interface. Making `fetch_hydstra_timeseries()` incorporate some of this is in development.

```{r, include=FALSE}
# For httptest2 mocking
end_vignette()
```
