---
title: "hydstra workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hydstra workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


library(httptest2)
start_vignette('hydstra_workflow')
```

```{r}
library(hydrogauge)
library(ggplot2)
```

At present, we largely assume the user will have a list of gauges they want, as the current ability to programatically obtain gauge numbers according to criteria is limited (but see `get_sites_by_datasource()`, which allows asking for all sites that have a given datasource). This vignette walks through the process with the core functions, for experimental wrappers that abstract some of this, see [their vignette](Hydstra-wrapper.Rmd).

To get timeseries, the user needs to ask for specific variables and timespans. Sometimes these are known *a priori*, e.g. if a gauge was chosen because it is known to have flow for a desired period. However, finding available variables and their periods of record can also be done through the functions here. This is one of the main purposes of this package; we want to be able to query available data.

This vignette will proceed with a set of sites chosen to span a range of characteristics useful for this demonstration.

-   The Upper Steavenson (405328) only has flow

-   Barwon (233217) has many variables, but their start dates differ

-   Taggerty (405331) is no longer in operation- ran 2010-2013

-   Marysville golf course (405837) is only rainfall

The functions all require gauges to be their numeric codes as characters. The API needs a comma-separated string (`"number1, number2"` ) , but the functions here will accept a vector `c("number1", "number2"`) and decompose it internally. This is typically easier and reflects more common R workflows such as having a column of site numbers in a dataframe.

```{r}
barwon <- '233217'
steavenson <- '405328'
taggerty <- '405331'
golf <- '405837'
```

# Querying available data

Before asking for timeseries data, we want to ask what data is available. we use  

## Finding datasources

To see what datasources are available for a site, use `get_datasources_by_site()`. I typically use "A", but it's worth looking to see what datasources are available for a target site(s), and then doing the next step (finding variables) for each, to see whether the available variables (or timeperiods) differ. *Note- there are often other datasources that work but are not returned here.*

```{r}
ds <- get_datasources_by_site(portal = 'Vic', 
                              site_list = c(barwon, steavenson, 
                                            taggerty, golf))
```

```{r rows.print = 15}
ds
```

Plot that to see data availability (\@ref(fig:datasource)).

```{r datasource, fig.cap = "Datasources available for each gauge. These are what are returned by the API, but may not be complete. Specifying other datasources on a pull may work."}
#| label: fig-datasource
#| fig-cap: Datasources available for each gauge. These are what are returned by the API, but may not be complete. Specifying other datasources on a pull may work.
plot_datasources_by_site(ds)
```

## Finding available variables and timespans

We then need to know what variables are available to extract timeseries of. We use `get_variable_list()` to get this information, including both their names and numbers, as well as other details such as the time period of record and units.

```{r}
var_info <- get_variable_list(portal = 'Vic', 
                              site_list = c(barwon, taggerty, 
                                            steavenson, golf), 
                              datasource = "A")
```

That returns a tibble with information about each gauge and variable (@tbl-vars). A few things to note- it gives the names of the gauges, the names and values of the variables, and a start and end date for each. For example, the Barwon's start date for stage (100) is 1961, while the others (pH, ppm, etc) didn't start until 2010.

*Note that this does **not** return derived discharge variables (140 and 141)*. If variable 100 (stage height) exists, the other two usually do, though sometimes not if there is no ratings curve.

```{r}
#| label: tbl-vars
#| tbl-cap: Variables available and period of record for target gauges
#| rows.print: 15 
var_info
```

Depending on the goals, it can be helpful to visualise this as the availability of each variable (\@ref(fig:vars-duration)) or the period of record of each variable (\@ref(fig:vars-period)).

```{r vars-duration, fig.cap = "Availability of each variable at each gauge, with color indicating the duration of record in days."}
#| label: fig-vars-duration
#| fig-cap: Availability of each variable at each gauge, with color indicating the duration of record in days.
var_info |> 
  dplyr::mutate(duration = period_end-period_start) |> 
ggplot(aes(x = var_name, y = site, fill = duration)) +
  geom_tile() +
  scale_fill_viridis_c(option = 'plasma') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r vars-period, fig.cap = "Availability of each variable at each gauge, with the period of record indicated by lines."}
#| label: fig-vars-period
#| fig-cap: Availability of each variable at each gauge, with the period of record indicated by lines.
var_info |> 
  tidyr::pivot_longer(cols = starts_with('period'), names_to = 'startend', values_to = 'date') |> 
ggplot(aes(y = date, x = site, color = var_name)) +
  geom_point(position = position_dodge(width = 0.5)) + geom_line(position = position_dodge(width = 0.5)) +
  coord_flip()
```

# Obtaining timeseries

This is typically the main goal, with the other steps getting us to the point of knowing what to ask for. Specifically, `get_variable_list()` gives us a reference to know what the variables are to ask for and the relevant timeperiods.

## Basic operation

In general, we use `get_ts_traces()` for a set of sites, variables, timeperiods, and statistics. The experimental wrapper functions `fetch_hydstra_timeseries()` and `fetch_timeseries()` call `get_ts_traces()` internally. In any case, there are some pitfalls to avoid.

If we just want a set of variables that all need the same statistic applied (e.g. daily mean flows), we can pass that in as a vector. For example, to get daily mean stage height (100), discharge (here in ML/day, 141), and temperature (450), we can do that in one call, even for multiple gauges. Asking here for one year to keep the call quick.

```{r}
ts_days <- get_ts_traces(portal = 'Vic', 
                         site_list = c(barwon, steavenson, taggerty, golf),
                         datasource = "A", 
                         var_list = c("100", "141", "450"),
                         start_time = 20200101,
                         end_time = 20201231,
                         interval = "day",
                         data_type = "mean",
                         multiplier = 1,
                         returnformat = 'df')
```

That returns a tall dataframe with both the requested values and some site metadata including the site name, location, etc (@tbl-ts), which the user can then split up or plot how they want (e.g. \@ref(fig:ts)). There are other options that return lists of dataframes if the user does not want all sites and variables combined-

-   `returnformat = "varlist"` a list with one tibble per variable

-   `returnformat = "sitelist"` a list with one tibble per site

-   `returnformat = "sxvlist"` a list with one tibble per site x variable combo (including empty lists for missing combos)

```{r tstab, tab.cap = "Head of returned timeseries"}
#| label: tbl-ts
#| tbl-cap: Head of returned timeseries
# rows.print doesn't really work with devtools::build_readme(), so use head
head(ts_days, 30)
```

```{r ts, fig.cap = "Timeseries of requested data, where available."}
#| label: fig-ts
#| fig-cap: Timeseries of requested data, where available.
ts_days |> 
  ggplot(aes(x = time, y = value, color = variable_short_name)) +
  geom_line() +
  facet_grid(variable_short_name ~ site_short_name, scales = 'free')
```


Note that if a variable isn't available for a gauge it just isn't returned, and same with timeperiods. We requested data from all four sites, but only the Barwon returns all variables. The golf course gauge does not return anything because it does not collect these variables, the Steavenson returns level and discharge but not temp, and the Taggerty doesn't appear at all despite having these variables because we've asked for data after it was decommissioned.

## Multiple variables, multiple statistics

Now, if we want another set of variables that should have a different statistic (e.g. rainfall makes sense as the daily sum, not the mean), we need a separate call to get_ts_traces() with a different `data_type` argument.

Note that again this will ignore gauges without the info (\@ref(tab:ts-rain-tab), \@ref(fig:ts-rain-fig)).

```{r}
ts_rain <- get_ts_traces(portal = 'Vic', 
                         site_list = c(barwon, golf), 
                         datasource = "A", 
                         var_list = c("10"),
                         start_time = 20200101,
                         end_time = 20201231,
                         interval = "day",
                         data_type = "tot",
                         multiplier = 1,
                         returnformat = 'df')
```

```{r ts-rain-tab, tab.cap="Head of returned rainfall timeseries"}
#| label: tbl-ts-rain
#| tbl-cap: Head of returned rainfall timeseries
head(ts_rain, 30)
```

```{r ts-rain-fig, fig.cap = "Timeseries of rainfall data, where available."}
#| label: fig-ts-rain
#| fig-cap: Timeseries of rainfall data, where available.
ts_rain |> 
  ggplot(aes(x = time, y = value, color = variable_short_name)) +
  geom_line() +
  facet_grid(variable_short_name ~ site_short_name, scales = 'free')
```

If the user wants to combine across different statistics, use `dplyr::bind_rows()` to combine post-hoc.

An automated approach that can simplify some common workflows (especially pulling period of record for many gauges) is available in `fetch_hydstra_timeseries()`, but care must be taken to avoid inappropriate statistics. See [the article](Hydstra-wrapper.Rmd).  


```{r, include=FALSE}
# For httptest2 mocking
end_vignette()
```
