---
title: "KiWIS workflow"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(hydrogauge)
library(ggplot2)
```

At present, we largely assume the user will have a list of gauges they want. The KiWIS interface does provide some opportunity to programatically find gauge numbers according to criteria, using `getParameterList()`, `getGroupList()`, and `getStationList()` (most useful). This vignette, however, assumes we have selected a set of gauges and want to identify variables and pull their timeseries. For experimental wrappers that abstract some of this, see [their vignette](kiwis-wrapper.Rmd).

Unlike Hydstra, which provides tailored arguments for different sorts of API control, KiWIS primarily uses text search within columns. This can be more flexible, but means we need to know column names and pay close attention to the regex used to filter those columns to avoid contaminating outputs. This generality means that no column is favoured over others in filtering, and hydrogauge provides access to the full search capability with the `extra_list` argument. However, we also make a concession for consistency by making `station_no` its own argument, corresponding to the `site_list` argument in Hydstra.  

# Querying available data

To get timeseries, the user needs to ask for specific variables and timespans. Sometimes these are known *a priori*, e.g. if a gauge was chosen because it is known to have flow for a desired period. However, finding available variables and their periods of record can also be done through the functions here, primarily `getTimeseriesList()`. This is one of the main purposes of this package; we want to be able to query available data.

## Finding available variables and timespans

Due to the search functionality of the KiWIS interface, we can use gauge numbers as we do with hydstra, but we can also search more generally. Note also that this returns a very large list, primarily due to the values in `ts_id` and `ts_name` columns, which arise because the various types of data and aggregations are given unique values there, rather than being calculated, as they are for hydstra.

```{r}
station_tslist <- getTimeseriesList(portal = 'bom', station_no = c('410730', 'A4260505'))

station_tslist
```

We can visualise the time period for each parameter at each gauge (here limited only to the QA'ed daily means).

```{r}
#| label: fig-vars-period
#| fig-cap: Availability of each variable at each gauge, with the period of record indicated by lines.
station_tslist |> 
  dplyr::filter(grepl('DMQaQc.Merged.DailyMean.24HR', ts_name)) |> 
  tidyr::pivot_longer(cols = c(from, to), names_to = 'startend', values_to = 'date') |> 
ggplot(aes(y = date, x = station_no, color = parametertype_name)) +
  geom_point(position = position_dodge(width = 0.5)) + 
  geom_line(position = position_dodge(width = 0.5)) +
  coord_flip()
```

We can take advantage of the search capability to ignore gauge numbers entirely, returning all sites meeting some regex pattern, here those with `'River Murray'` in the `station_name`. We do this with the `extra_list` argument, which takes column names as names and the seach pattern as the item. This approach works for any column, and so we also only look at the "DMQaQc" data at a Daily Mean aggregation. The `returnfields` argument lets us choose which columns to return. We include 'coverage' here to the returnfields to get the period of record.


```{r}
RM_ts <- getTimeseriesList(portal = 'bom',
                           extra_list = list(station_name = 'River Murray*',
                                             ts_name = 'DMQaQc.Merged.DailyMean.24HR'),
                           returnfields = c('station_no', 'station_name',
                                            'ts_name', 'ts_id', 
                                            'ts_unitname', 'parametertype_name',
                                            'coverage'))

RM_ts
```

We can visualise the availability of each variable at each gauge (\@ref(fig:vars-duration)).

```{r vars-duration, fig.cap = "Availability of each variable at each gauge, with color indicating the duration of record in days."}
#| label: fig-vars-duration
#| fig-cap: Availability of each variable at each gauge, with color indicating the duration of record in days.
RM_ts |> 
  dplyr::mutate(duration = to-from) |> 
ggplot(aes(x = parametertype_name, y = station_no, fill = duration)) +
  geom_tile() +
  scale_fill_viridis_c(option = 'plasma') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.callout-note}

The available returnfields are poorly documented. The default are those returned above for `station_tslist`, 

```{r}
names(station_tslist)[!names(station_tslist) %in% c('from', 'to', 'database_timezone')]
```
But there are others that can be requested. In particular, 'coverage' is needed to get the period of record.

```{r}
  # According to kisters, these exist
  all_return <- c('station_name', 'station_latitude', 'station_longitude', 'station_carteasting', 'station_cartnorthing', 'station_local_x', 'station_local_y', 'station_georefsystem', 'station_longname', 'ts_id', 'ts_name', 'ts_shortname', 'ts_path', 'ts_type_id', 'ts_type_name', 'parametertype_id', 'parametertype_name', 'stationparameter_name', 'stationparameter_no', 'stationparameter_longname', 'ts_unitname', 'ts_unitsymbol', 'ts_unitname_abs', 'ts_unitsymbol_abs', 'site_no', 'site_id', 'site_name', 'catchment_no', 'catchment_id', 'catchment_name', 'coverage', 'ts_density', 'ts_exchange', 'ts_spacing', 'ts_clientvalue##', 'datacart', 'ca_site', 'ca_sta', 'ca_par', 'ca_ts')
  # I get http 500 errors unless cut to
  sub_return <- all_return[c(1:34, 37:40)]
  sub_return
```

:::

# Obtaining timeseries

To pull timeseries, we need to know either the `ts_id` or `ts_path` arguments that we want. This filters on those columns to ensure we get the timeseries we want. Note in both `station_tslist` and `RM_ts` that each row (defined by station, variable, aggregation, QA'd, etc) gets its own `ts_id` value. We need to choose the ones we want.

:::{.callout-tip}

The find_ts_id() function helps us find desired `ts_id` values. It is a wrapper over `getTimeseriesList()` with regex to select what we want. As such, it's not in the raw API workflow, but is very handy. See [the wrapper vignette](kiwis-wrapper.Rmd)

:::

:::{.callout-note}

In choosing these `ts_ids`, we're making many of the same decisions we would make when asking the hydstra function `get_ts_traces()` for `site_list`, `datasource`, `var_list`, `interval`, `data_type`, and `multiplier`. This has pluses and minuses. It's much harder here to ask for what we want (but see `find_ts_ids()`), but because each aggregation is pre-supplied and indexed uniquely it's much easier to get different aggregations from different variables in one call.

:::

Let's demonstrate with some ts_id values from `station_tslist` to pull daily mean QaQc'ed values for level, discharge, and water temp. Note that we have to ask for discharge and level with separate ts_ids for each gauge. This is another argument for using `find_ts_id()`.

```{r}
ts_example <- c(
  '208669010', '208648010', # level and discharge Lock 9
  '1573010', '1598010', # level and discharge Cotter R.
  '380167010'
  )
```

We'll just pull one year

```{r}
ts_example <- getTimeseriesValues(portal = 'bom',
                                  ts_id = ts_example,
                                  start_time = 20100101,
                                  end_time = 20101231)
```

That provides a tall dataframe containing additional information about the gauge (@tbl-ts). The returned columns can again be adjusted with `returnfields` and `meta_returnfields`.

```{r}
#| label: tbl-ts
#| tbl-cap: Head of returned timeseries
# rows.print doesn't really work with devtools::build_readme(), so use head
head(ts_example, 30)
```

```{r}
#| label: fig-ts
#| fig-cap: Timeseries of requested data, where available.
ts_example |> 
  ggplot(aes(x = time, y = value, color = parametertype_name)) +
  geom_line() +
  facet_grid(parametertype_name ~ station_name, scales = 'free', labeller = label_wrap_gen(10))
```

A common use will be asking for the period of record, so we show that here for these two gauges. We use the `period = 'complete'` argument to take advantage of internal API functionality (which also allows period units like 'P2W').

```{r}

ts_all <- getTimeseriesValues(portal = 'bom',
                                  ts_id = c('208648010', '1573010'),
                                  period = 'complete')

```

```{r}
#| label: fig-ts-all
#| fig-cap: Timeseries of period of record for discharge.
ts_all |> 
  # dplyr::filter(value >= 0) |> 
  ggplot(aes(x = time, y = value, color = station_name)) +
  geom_line() +
  facet_grid(station_name ~ ., scales = 'free', labeller = label_wrap_gen(10)) +
  theme(legend.position = 'none')
```




An automated approach that can simplify some common workflows (especially pulling period of record for many gauges and programatically selecting variables across gauges) is available in `fetch_kiwis_timeseries()`. See [the article](kiwis-wrapper.Rmd). 
